### 2017-07-26 chapters/introduction/mechanics.tex
Write summary. Only missing to write about fenics and summary of  the unpublished manuscripts.
Write about fenics

The active stress approach is based one the classical three element
Hill model illustrated in Figure \ref{fig:hill_muscle_model}, where the
active contribution naturally decomposes the total stress into a sum
of passive and active stresses
\cite{nash2004electromechanical}. Hence, in the active stress
formulation \cite{hunter1998modelling} one assumes that the total
Cauchy stress $\Cauchy$ can be written as an additive sum of one
passive contribution $\Cauchy_p$ and one active contribution $\Cauchy_a$,
\begin{align}
  \Cauchy = \Cauchy_p + \Cauchy_a && J \Cauchy_p =  \frac{\partial \Psi(\F)}{\partial \F} \F^{T}.
\end{align}
The passive contribution is determined by the material model used,
while the active contribution is given by 
\begin{align}
  \Cauchy_a = \Cauchy_{ff} \Fef \otimes \Fef +
  \Cauchy_{ss} \mathbf{s} \otimes \mathbf{s} +
  \Cauchy_{nn} \mathbf{n} \otimes \mathbf{n},
\end{align}
and the different constants $\Cauchy_{ff}, \Cauchy_{ss}$, and
$\Cauchy_{nn}$, which are the active stress in the fiber, sheet and
sheet-normal direction respectively, are typically coupled to the
electrophysiology and calcium dynamics.
There are experimental evidence that the active stresses in the
transverse direction of the fibers ($\Cauchy_{ss}$, and $\Cauchy_{nn}$),
are non-negligible \cite{lin1998multiaxial}, and one approach is to assume
a uniform transverse activation in which the total active tension
can be written as 
\begin{align}
  \Cauchy_a = T_a \left[\Fef \otimes \Fef +
   \eta\left( \mathbf{s} \otimes \mathbf{s} +
  \ \mathbf{n} \otimes \mathbf{n} \right)\right],
\end{align}
where $\eta$ represent the amount of transverse activation and $T_a
\in \mathbb{R}$ is the magnitude of the active tension.
In the limiting case ($\eta = 0.0$), the active tension acts purely
along the fibers and is reduced to 
\begin{align}
  \Cauchy_a = T_a \Fef \otimes \Fef.
\end{align}
Note that, by observing  that
\begin{align*}
  \frac{\partial I_{4\mathbf{a}_0}}{\partial \F} = \frac{\partial (\mathbf{a}_0  \cdot \C \mathbf{a}_0 )}{\partial \F}
  = 2 \mathbf{a} \otimes \mathbf{a}_0 \implies  \mathbf{a} \otimes  \mathbf{a}= \frac{1}{2} \frac{\partial I_{4\mathbf{a}_0}}{\partial \F} \F^{T}
\end{align*}
and that $I_1 =  I_{4\ef} +  I_{4\eS} +  I_{4\en}$, 
we can instead decompose the strain-energy function into a passive and active
parts \cite{pathmanathan2010cardiac}, $\Psi= \Psi_p + \Psi_a$, with
\begin{align}
\Psi_a = \frac{T_a}{2J} \left(( I_{4\ef} - 1)  + \eta \left[ (I_1 - 3) -
    (I_{4\ef} - 1)\right] \right), 
\end{align}
so that $J \Cauchy_a  = \frac{\partial \Psi_a}{\partial \F}
\F^{T}$.

@@

We add the constraint $\det(\F_a) = 1$, meaning that the active
deformation is volume preserving. Further we assume that the activation is
transversely isotropic, so that the sheet and sheet-normal axis is
treated in the same way. It is then straight forward to verify that
$\gamma_n = \gamma_s =1- (1-\gamma_f)^{-1/2}$, and we have
\begin{equation}
  \F_a = (1 - \gamma) \ef \otimes \ef  + \frac{1}{\sqrt{1 - \gamma}} (\I - \ef \otimes \ef), 
 \label{eq:active_strain_Fa_gjerald}
\end{equation}
where we set $\gamma = \gamma_f$ for convenience.

@@

\subsection{Implementation details}
The cardiac mechanics solver developed during the work of this thesis
is implemented using the finite element framework FEniCS. Here we
briefly explain the main components of FEniCS as well as some
numerical considerations made when implementing the solver.
For more information about FEniCS, the reader is referred to the
official web page (\url{https://fenicsproject.org}).

@@

\subsubsection{FEniCS}
\label{sec:fenics}

FEniCS is a framework for solving partial differential
equations using the finite element method, and consist of the
following software components:

FEniCS has a python
interface where the user can write code which is close to the
mathematical notions used in the theory.

code written in
FEniCS is close to the mathematical notation, and the handling of
assembly code happens ``under the hood'', which makes it easy to use
and intuitive for scientists

>>>>

The active stress approach is based one the classical three element
Hill model illustrated in Figure \ref{fig:hill_muscle_model}, where the
active contribution naturally decomposes the total stress into a sum
of passive and active stresses
\cite{nash2004electromechanical}. Hence, in the active stress
formulation \cite{hunter1998modelling} one assumes that the total
Cauchy stress $\Cauchy$ can be written as an additive sum of one
passive contribution $\Cauchy_p$ and one active contribution $\Cauchy_a$,
\begin{align}
  \Cauchy = \Cauchy_p + \Cauchy_a && J \Cauchy_p =  \frac{\partial \Psi(\F)}{\partial \F} \F^{T}.
\end{align}
The passive contribution is determined by the material model used,
while the active contribution is given by 
\begin{align}
  \Cauchy_a = \Cauchy_{ff} \Fef \otimes \Fef +
  \Cauchy_{ss} \mathbf{s} \otimes \mathbf{s} +
  \Cauchy_{nn} \mathbf{n} \otimes \mathbf{n},
\end{align}
and the different constants $\Cauchy_{ff}, \Cauchy_{ss}$, and
$\Cauchy_{nn}$, which are the active stress in the fiber, sheet and
sheet-normal direction respectively, are typically coupled to the
electrophysiology and calcium dynamics.
There are experimental evidence that the active stresses in the
transverse direction of the fibers ($\Cauchy_{ss}$, and $\Cauchy_{nn}$),
are non-negligible \cite{lin1998multiaxial}, and one approach is to assume
a uniform transverse activation in which the total active tension
can be written as 
\begin{align}
  \Cauchy_a = T_a \left[\Fef \otimes \Fef +
   \eta\left( \mathbf{s} \otimes \mathbf{s} +
  \ \mathbf{n} \otimes \mathbf{n} \right)\right],
  \label{eq:intro_active_stress}
\end{align}
where $\eta$ represent the amount of transverse activation and $T_a
\in \mathbb{R}$ is the magnitude of the active tension.
In the limiting case ($\eta = 0.0$), the active tension acts purely
along the fibers and is reduced to 
\begin{align}
  \Cauchy_a = T_a \Fef \otimes \Fef.
\end{align}
Note that, by observing  that
\begin{align*}
  \frac{\partial I_{4\mathbf{a}_0}}{\partial \F} = \frac{\partial (\mathbf{a}_0  \cdot \C \mathbf{a}_0 )}{\partial \F}
  = 2 \mathbf{a} \otimes \mathbf{a}_0 \implies  \mathbf{a} \otimes  \mathbf{a}= \frac{1}{2} \frac{\partial I_{4\mathbf{a}_0}}{\partial \F} \F^{T}
\end{align*}
and that $I_1 =  I_{4\ef} +  I_{4\eS} +  I_{4\en}$, 
we can instead decompose the strain-energy function into a passive and active
parts \cite{pathmanathan2010cardiac}, $\Psi= \Psi_p + \Psi_a$, with
\begin{align}
\Psi_a = \frac{T_a}{2J} \left(( I_{4\ef} - 1)  + \eta \left[ (I_1 - 3) -
    (I_{4\ef} - 1)\right] \right), 
\end{align}
so that $J \Cauchy_a  = \frac{\partial \Psi_a}{\partial \F}
\F^{T}$.

@@

We add the constraint $\det(\F_a) = 1$, meaning that the active
deformation is volume preserving. Further we assume that the activation is
transversely isotropic, so that the sheet and sheet-normal axis is
treated in the same way. It is then straight forward to verify that
$\gamma_n = \gamma_s =1- (1-\gamma_f)^{-1/2}$, and we have
\begin{equation}
  \F_a = (1 - \gamma) \ef \otimes \ef  + \frac{1}{\sqrt{1 - \gamma}} (\I - \ef \otimes \ef), 
 \label{eq:intro_active_strain_Fa_gjerald}
\end{equation}
where we set $\gamma = \gamma_f$ for convenience.

@@

\subsection{Implementation details}
The cardiac mechanics solver developed during the work of this thesis
is implemented using the finite element framework FEniCS. Here we
briefly explain the main components of FEniCS as well as some
numerical considerations made when implementing the solver.

@@

\subsubsection{The FEniCS Project}
\label{sec:fenics}

The FEniCS project is an open-source computing platform for solving
partial differential equations (PDEs) using the finite element method (FEM).
Solving PDEs using FEM involves many implementation details that can
be tedious to implement yourselves. The idea behind FEniCS is to
automate code generation so that the user can spend more time on doing
research and less time on implementation of assembly matrices. At the
core of FEniCS is DOLFIN \cite{logg2012dolfin}, which is C++/Python
library, and works as the main interface in FEniCS. In this thesis
only the Python interface has been used, in which C++ code is
automatically generated using SWIG. This allows for simplicity through
the Python scripting language and the speed of the C++ language.
The domain specific language used to represent weak formulations is
called the Unified form language (UFL) \cite{alnaes2014unified}, and
allows for e.g automatic differentiation of forms and expressions. The
FEniCS form compiler (FFC) \cite{logg2012ffc} compiles code written in
UFL to Unified Form-assembly Code (UFC) \cite{alnaes2012ufc} which are
optimized C++ code. The Python interface also makes use of the Instant
module which allows for just-in-time (JIT) compilation of C++
code. The compiled code is also stored in a cache so that compilation
of a form only happens once. Also, the relatively new UFL Analyser and
Compiler System (UFLACS) allows for fast compilation of complex forms
such as forms such as variational formulations that include the Holzapfel
Ogden material model \eqref{eq:holzapel_full}.

For more information about FEniCS, the reader is referred to the
official web page (\url{https://fenicsproject.org}) or any of the
cited literature.
